---
title: "lofi-colour"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{lofi-colour}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

suppressPackageStartupMessages({
  library(dplyr)
  library(purrr)
  library(lofi)
})
```

```{r setup}
library(lofi)
```


# Introduction

Colours in R are usually stored as 8-bits per channel for red, green and blue - 
with a total of 24-bits then required for a single RGB colour.  This means that 
there are a total of `2^24 = 16777216` possible colour values.

Each 8-bit value is usually represented as hex e.g. `255 = FF`, and then all 
the values are concatenated to specify a colour e.g.`red = #FF0000`.

Note: `lofi` has no built-in support for colours with alpha values.

# Storing a colour in 8-bits

To store a colour in 8-bits, `lofi` truncates the lower bits from each of the 
colour channels, keepingly 3 bits for red and green channels, and only 2 bits
for blue.  More information about the 8-bit colour format is [available on wikipedia](https://en.wikipedia.org/wiki/8-bit_color).


8-bit colour means that only 256 colours can be represented, so conversion to 
this low-fidelity representation and back again will results in the loss of
colour accuracy.  


# Rainbow palette in 8-bits

In the following example, 8 colours from a rainbow palette are converted to 
8-bit colour and then reconstructed.  Visually, the reconstructed palette looks 
very close to the original colours.

```{r, fig.height = 3}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Create a rainbow palette
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
(pal <- rainbow(8))
lofi::plot_palette(pal)
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Convert to 8-bit lofi representation and back again
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
lofi_pal <- lofi::hex_colour_to_lofi(pal, rgb_bits = c(3, 3, 2))
reconstructed <- lofi::lofi_to_hex_colour(lofi_pal, c(3, 3, 2))
lofi::plot_palette(reconstructed)
```





# Viridis palette in 8-bits

In the following example, 64 colours from a viridis palette are converted to 
8-bit colour and then reconstructed.  Visually, the reconstructed palette looks 
similar to the original colours, but the errors in colour accuracy are quite apparent.

```{r, fig.height = 3}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Create a viridis palette
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
pal <- viridisLite::plasma(64)
lofi::plot_palette(pal, width = 100, height = 100, ncol = 8, xf = 0.95)
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Convert to 8-bit lofi representation and back again
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
lofi_pal <- lofi::hex_colour_to_lofi(pal, rgb_bits = c(3, 3, 2))
reconstructed <- lofi::lofi_to_hex_colour(lofi_pal, c(3, 3, 2))
lofi::plot_palette(reconstructed, width = 100, height = 100, ncol = 8, xf = 0.95)
```




# Viridis palette in 12-bits

In the following example, 64 colours from a viridis palette are converted to 
12-bit colour and then reconstructed.  Visually, the reconstructed palette looks 
similar to the original colours, and the errors in colour accuracy are less apparent
than with only 8-bits

```{r, fig.height = 3}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Convert to 12-bit lofi representation and back again
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
lofi_pal <- lofi::hex_colour_to_lofi(pal, rgb_bits = c(4, 4, 4))
reconstructed <- lofi::lofi_to_hex_colour(lofi_pal, c(4, 4, 4))
lofi::plot_palette(reconstructed, width = 100, height = 100, ncol = 8, xf = 0.95)
```




















