---
title: "packing-specification"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{packing-specification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(lofi)
```


Pack/Unpack values as low-fidelity representation into a single 32-bit integer.
These functions require a packing spec (\code{pack_spec}) which defines how values
are converted to/from their low-fidelity representations.

##  Packing Specification Overview:

The packing specification (`pack_spec`) is a named list detailing how
values should be converted to their low-fidelity representations.

The name of the values in the `pack_spec` correspond to the names in
the `values` argument to `pack()`.

The following are valid \code{types} for packing:

* integer - A standard integer value
* double - A standard double precision floating point value
* logical - A standard R logical values
* choice - Analogous to storing a factor
* scaled - Storing a scaled value in the available bits and then
            re-scaling when unpacking
* colour - A standard hex colour value of the form '#123456'
* custom - User must specify functions to convert value to
            low-fidelity representation and reconstruction
 


##  Packing Specification - Integer

Integers are packed by truncating leading bits that aren't needed e.g. the number 8
only needs 4 bis to represent it, and the other leading 28 bits can be ignored.

Packing an integer in this way is lossless - the reconstructed value using
`unpack() will be identical to the original value.

* `nbits` - total number of bits to use
* `signed` - keep a sign bit? Default: FALSE
* `mult` - pre-scale the value when packing, and undo scaling
when unpacking. i.e. `(value + offset) * mult. Default: 1
* `offset` - offset the value when packing, and undo offset when
unpacking . i.e. `(value + offset) * mult. Default: 0




##  Packing Specification - Double

Doubles are packed by truncating the mantissa and re-encoding the exponent.
This will almost definitely lead to loss of precision, and any reconstructed
value will not be identical to the original.

* `float_name` - name of floating point representation to use. Options:
    * 'single' - single precision 32-bits
    * 'fp24' - 24bit float
    * 'bfloat16' - 16 bit float
    * 'half' - half precision 16-bits
* `nbits` - total number of bits to use. If specified, this takes
   precedence over the `float_name value.
* `maxval` - only used if `nbits is specified. Used to
  calculate the total bits in the exponent.
* `signed` - keep a sign bit? Only used if `nbits is
specified. Default: FALSE
* `float_bits` - [advanced] A 3-element numeric vector giving the
  number of bits to assign to the sign, exponent and mantissa,
  respecively.  If given, `float_bits`  takes precedence over
  both `nbits` and `float_name`.



##  Packing Specification - Logical

Logical values only require a single bit, but more bits can be specified
if desired.

* `nbits` - total number of bits to use. Optional. Default: 1



##  Packing Specification - Choice

A choice is very similar to a factor, but the labels are only stored in the
specification, and the index is 0-based (instead of 1-based)

* `options` - vector of options to match values against.  Only
  the index of the value into this list is stored.
* `nbits` - total number of bits to use. Optional. If not given,
  it is calculated to be the number of bits necessary to store all
  possible options.
 


##  Packing Specification - Scaled

Specifying the stored of a scaled value is sometimes easier than trying to
work out how to corectly store a double precision floating point.

* `nbits` - number of bits to use
* `min` - minimum value to be stored. Default: 0
* `max` - maximum value to be stored in the given bits. Every
stored value is scaled by `(2^nbits - 1)/max
when `pack()ed`, and unscaled when `unpacked()`.


##  Packing Specification - Colour

Packing a colour is achieved by truncating the bits for each of the R, G
and B channels separately.

* `nbits` - number of bits to use
* `rgb_bits` - [advanced] A 3-element numeric vector giving the
number of bits to assign to the R, G and B channels
respecively.  If given, `rgb_bits` takes precedence over
`nbits`.
 


##  Packing Specification - Custom

This packing specification allows you to specify a function to convert a
value into a low-fidelity representation, and a matching function to
reconstruct the original value.

* `nbits` - number of bits to use
* `pack_func` - a function. Alternatively, can specify a formula
  which will be converted internally to a function taking a single
  argument,  `.x`.
* `unpack_func` - a function. Alternatively, can specify a formula
  which will be converted internally to a function taking a single
  argument,  `.x`.



# Example 1

```{r}
 # Specify how to convert values to low-fidelity representation
 pack_spec <- list(
   x      = list(type = 'double', float_name = 'bfloat16'),
   valid  = list(type = 'logical'),
   stars  = list(type = 'integer', mult = 10, nbits = 6),
   alpha  = list(type = 'scaled', max = 1, nbits = 5),
   grade  = list(type = 'choice', options = c('A', 'B', 'C', 'D'))
 )

 # Assemble the values
 values <- list(
   x      = 1.234e21,
   valid  = TRUE,
   stars  = 4.5,     # Star rating 0-5. 1 decimal place.
   alpha  = 0.8,     # alpha. range [0, 1]
   grade  = 'B'
 )

 # pack them into a single integer
 packed_values <- pack(values, pack_spec)

 # Reconstruct the initial values from the packed values
 unpack(packed_values, pack_spec)
```




# Example 2 - Custom packing

In the following example, the same distance value is encoded in two different ways.

The first method simply encodes the value as a 16 bit floating point value known as a 
[bfloat16](https://en.wikipedia.org/wiki/Bfloat16_floating-point_format).

In the second method the `pack_func` is used to take the log of the value, 
multiply it by 100 and then store this as a 16bit integer.  The `unpack_func`
defines the reverse.


```{r}
pack_spec <- list(
  distance1 = list(type = 'double', float_name = 'bfloat16'),
  distance2 = list(type = 'custom', nbits = 16,
                   pack_func   = ~100 * log(.x), unpack_func = ~exp(.x/100))
)

values <- list(
  distance1 = 1.234e21,
  distance2 = 1.234e21
)

packed_values <- pack(values, pack_spec)

unpack(packed_values, pack_spec)
```




# Example 3 - Custom packing of a quantized angle

In the following example, an angle in the range [0, 180] is stored in just 
2 bits by quantizing it to the nearest 45 degrees.


```{r}
pack_spec <- list(
  angle = list(
    type        = 'custom',
    nbits       =  2,
    pack_func   = ~round((.x %% (3.5/4*180)) / 45),
    unpack_func = ~.x * 45
  )
)

values <- list(
  angle = 60
)

packed_values <- pack(values, pack_spec)

unpack(packed_values, pack_spec)
```




